---
title: "Bio201 Final Project"
author: "Yingyue Ding","Kevin Huang","Sunyimeng Lei", "Clarissa Halim" 
date: "11/18/2019"
output: html_document
---

Question: 
How does the consumption of a prebiotic effect the gut microbiome?
SCFAs, pH, Gases 
Microbial community (diversity, changes in taxon abundance

??????

Lmaoooooooo

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "~/Documents/Bio201ThursPM/")
```

# Load packages
```{r Load packages, eval=TRUE, echo=TRUE, message=FALSE, warning=FALSE, results='hide'}
library(tidyverse)
library(readxl)
library(broom)
library(cowplot)
library(phyloseq)
library(vegan)
set.seed(7)
```

# Import data

If not already loaded into the global environment, import the sequence variant, taxa, and participant data files from Lab 7. We use the same information for calculating both alpha and beta diversity metrics.

```{r}
# custom function for rounding
myfunction <- function(x) {
  x = round ( x * 1000 , digits = 0 )
}

# sequence variant information 
sv_table <- read_delim("~/Documents/Bio201ThursPM/final_project_data/shared_wkly.txt",
                            delim = ",", escape_double = FALSE, 
                            trim_ws = TRUE, na=c("NA"),
                            col_types = list()) %>%
  rename(sample_id = X1) %>% #add missing column name 
  mutate(sample_id = str_replace(sample_id, 
                                 pattern = "-", 
                                 replacement = "_")) %>% #replace hyphens with underscores in sample id column 
  mutate_if(is.numeric, myfunction) %>% 
  column_to_rownames(var = "sample_id") %>% #sample IDs need to be made into row names
  as.matrix() %>% #convert data frame to matrix object
  otu_table(., taxa_are_rows = FALSE) #specify type of phyloseq object

# taxanimc classifications 
taxa <- read_delim("Lab7/raw_data/seqvar_classifications.csv",
                            delim = ",", escape_double = FALSE, 
                            trim_ws = TRUE, na=c("NA"),
                            col_types = list()) %>%
  column_to_rownames(var = "sv") %>% #sequence variants (OTUs) need to be made into row names 
  as.matrix() %>%
  tax_table(.)  #specify type of phyloseq object

# participant and sample information 
all_data_indv <- read_delim("Lab7/raw_data/all_data_indv.txt", 
                            delim = "\t", escape_double = FALSE, 
                            trim_ws = TRUE, 
                            col_types = cols(Bristol_numeric = col_number(),
                                             Bristol_score = col_character(), 
                                             pH = col_number())) %>%
  rename_all(tolower) %>% #make all column names lower case
  distinct(., sample_id, .keep_all = TRUE) %>% #remove duplicate sample ids 
  column_to_rownames(var = "sample_id") %>% #sample IDs need to be made into row names
  sample_data(.) #specify type of phyloseq object

# merge all three data frames into phyloseq object
physq_obj <- phyloseq(sv_table, taxa, all_data_indv) 
physq_obj 
```

### Format data

As in lab 7 we are not using the Winter 2015 data, and we are only interested in semesters data from week 1 and week 3. We will make a subset of the phyloseq object then use that going forward. 
```{r}
physq_sub <- physq_obj %>%
  subset_samples(semester != "Winter2015") %>%
  subset_samples(quantity_compliant == "yes") %>%
  subset_samples(study_week == "week1" | study_week == "week3")
physq_sub
```

# Evenness

To determine if a community is even or uneven, we use stacked bar plots as a relative measure. Before generating the plot, we have to transform our abundance values from sequence counts to a percentage.

### Data transformation

The first step of the transformation process involves summarizing taxa at whatever rank (phylum through species) we plan to plot the data. We know from the previous code block we have 1349 taxa. We will agglomerate (sum) the counts for ESVs which are members of the same phyla. View the phyloseq object to determine the number of taxa; note how it is reduced compared to the starting data (1349 vs 12).
```{r}
tmp1 <- physq_sub %>%
  tax_glom(taxrank = "phylum")
tmp1
```

After we have our taxa aggregated at the taxanomic level of interest we need to convert the sequence counts to relative abundance. Relative abundance is represented in the literature sometimes as a percent (1-100%) or as a ratio (0-1). Either is acceptable, they convey the same information, it really is a matter of personal preference. Below we will write an in-line custom function to calculate relative abundance. Then we will use some phyloseq functions to view the transformed counts of our data to make sure the function behaved as expected. 
```{r}
tmp2 <- tmp1 %>%
  transform_sample_counts(function(x) { (x/sum(x)) * 100 } )
tmp2

# otu_table() extracts the counts
# as.data.frame() converts back to data frame class
# head() gives a preview of the data frame
head(as.data.frame(otu_table(tmp2))) 
```

To improve computer speed and performance remove any large objects from the global environment you will not use downstream. 
```{r}
rm(tmp1)
rm(tmp2)
```

As you know ggplot requires a long-format data frame. In the code blocks above we are working with phyloseq data, which is matrix-derived and actually its own class of object (i.e., not a data frame!). The phyloseq package contains a function, psmelt(), that takes a phyloseq object and converts it to a long-format data frame for us. We are going to pipe that function after the glom and transformation functions to create a data frame that is compatable with ggplot(). Be careful when using the psmelt() function, the new columns created do not follow the same column name style conventions we use in this course. 
```{r}
physq_long <- physq_sub %>%
  tax_glom(taxrank = "phylum") %>% #agglomerate at phylum level               
  transform_sample_counts(function(x) {x/sum(x)} ) %>% #calculate rel abund
  psmelt() %>% #melt to long format for ggplot
  rename_all(tolower) %>% #update new columns to match style conventions 
  select(-sample_number, -quantity_compliant, -use_data, 
         -ends_with("_mm"), -ends_with("_mmol_kg"), 
         -starts_with("bristol"), -ph) %>% #drop columns not using
  filter(abundance > 0.001) %>% #remove low abund taxa
  arrange(phylum) #sort alphabetically by taxa
```

### Generate plot 

Now we are ready to make our plot. The same prinicples apply, we are just using a new ggplot function, geom_col(). 
```{r}
ggplot(data = physq_long, 
       aes(x = study_week, #specify the arguments for each axis
           y = abundance, 
           fill = phylum)) + #set fill to the taxa rank used for glom
  geom_col() + #plot type for bars 
  facet_grid(~semester) + #panel for each semester
  xlab(NULL) + 
  theme(axis.text.x = element_text(angle = 90,
                                   vjust = 0.5)) + #clean up x-axis labels 
  ylab("Relative Abundance (Phyla > 0.1%)")
```

The plot above shows the relative abundace of phyla in participants who consumed BRMPS. Notice the legend has some odd formatting. We can use another ggplot function, scale_fill_manual(), to improve the readability, we will also use the scale_fill_manual() function to update the colors of the plot.  

Next week when you begin data analysis for the group project, we will use a consistent color for each taxa. To do this we have to manually change the colors in the plots; create a vector from hex codes for colors, then call that vector in scale_fill_manual() or scale_color_manual() to update the default ggplot colors. An example of this concept is show below. 
```{r}
# vector of color hex codes 
taxa_colors <- c('#ff7f7f','#33b5e5',
                '#ffbb33','#aa66cc',
                '#ff3298', '#99cc00',
                '#cc0000','#007299',
                '#cc6c00','#6b238e',
                '#b20058', '#436500',
                '#ff4444','#0099cc',
                '#ff8800','#9933cc',
                '#e50072', '#669900',
                '#7f0000','#004c66',
                '#995100','#3c1451',
                '#660033', '#2c4c00',
                '#8ed5f0','#ffb2b2',
                '#bc93d1','#ffe564',
                '#d2fe4c','#ff7fbf')
```
This following website is a useful color palette generating tool: https://medialab.github.io/iwanthue/ 

The crucial arguments to note in scale_file_manual() are breaks and labels. The arguments given to 'breaks' are the *current* legend labels. The arguments given to 'labels' are the *new* legend labels. Be careful when providing the arugments to labels, as they will be used to rename in the order typed. If you swap the order as you are typing in the arguments, the new labels in your plot will be incorrect. Updated code for the plot: 
```{r}
ggplot(data = physq_long, 
       aes(x = study_week, #specify the arguments for each axis
           y = abundance, 
           fill = phylum)) + #set fill to the taxa rank
  geom_col() + #plot type for bars 
  facet_grid(~semester) + #panel for each semester
  xlab(NULL) + 
  theme(axis.text.x = element_text(angle = 90,
                                   vjust = 0.5)) + #clean up x-axis labels 
  ylab("Relative Abundance (Phyla > 0.1%)") + #improve y-axis labels 
  # improve plot legend readability 
  scale_fill_manual(name = "Phyla", #change name for legend
                    values = taxa_colors, #indicate colors from vector of custom colors 
                    #current legend labels
                    breaks = c("Actinobacteria <actinobacteria>",
                              "Bacteroidetes <Bacteroidetes>",
                              "Cyanobacteria/Proteobacteria",
                              "Euryarchaeota",
                              "Firmicutes",
                              "Firmicutes/Actinobacteria_<actinobacteria>",
                              "Fusobacteria <Fusobacteria>",
                              "Lentisphaerae",
                              "Proteobacteria",
                              "Tenericutes",
                              "Verrucomicrobia"),
                    #new legend labels 
                    labels = c("Actinobacteria",
                              "Bacteroidetes",
                              "Cyanobacteria/Proteobacteria",
                              "Euryarchaeota",
                              "Firmicutes",
                              "Firmicutes/Actinobacteria",
                              "Fusobacteria",
                              "Lentisphaerae",
                              "Proteobacteria",
                              "Tenericutes",
                              "Verrucomicrobia"))
```

### Homework 8.1 
Generate a bar plot at the genus level, create and save the plot to a figures folder. 
```{r}
physq_long_genus <- physq_sub %>%
  tax_glom(taxrank = "genus") %>%             
  transform_sample_counts(function(x) {x/sum(x)} ) %>% 
  psmelt() %>% 
  rename_all(tolower) %>% 
  select(-sample_number, -quantity_compliant, -use_data, 
         -ends_with("_mm"), -ends_with("_mmol_kg"), 
         -starts_with("bristol"), -ph) %>% 
  filter(abundance > 0.1) %>% 
  arrange(genus)

ggplot(data = physq_long_genus, 
       aes(x = study_week, #specify the arguments for each axis
           y = abundance, 
           fill = genus)) + 
  geom_col() + 
  facet_grid(~semester) + 
  xlab(NULL) + 
  theme(axis.text.x = element_text(angle = 90,
                                   vjust = 0.5)) + 
  ylab("Relative Abundance (genus > 10%)")



```

### Homework 8.2
Generate a bar plot at the order level, create and save the plot to a figures folder. 
```{r}

physq_long_order <- physq_sub %>%
  tax_glom(taxrank = "order") %>% #agglomerate at phylum level               
  transform_sample_counts(function(x) {x/sum(x)} ) %>% #calculate rel abund
  psmelt() %>% #melt to long format for ggplot
  rename_all(tolower) %>% #update new columns to match style conventions 
  select(-sample_number, -quantity_compliant, -use_data, 
         -ends_with("_mm"), -ends_with("_mmol_kg"), 
         -starts_with("bristol"), -ph) %>% #drop columns not using
  filter(abundance > 0.001) %>% #remove low abund taxa
  arrange(order)

ggplot(data = physq_long_order, 
       aes(x = study_week, #specify the arguments for each axis
           y = abundance, 
           fill = order)) + #set fill to the taxa rank used for glom
  geom_col() + #plot type for bars 
  facet_grid(~semester) + #panel for each semester
  xlab(NULL) + 
  theme(axis.text.x = element_text(angle = 90,
                                   vjust = 0.5)) + #clean up x-axis labels 
  ylab("Relative Abundance (order > 0.1%)")
```


# Beta diveristy

### Conduct ordination analysis

This step calculates the diveristy index (using Bray Curtis in the example below), and does the ranking process of the diveristy indices to generate the information necessary to create a two-dimensional plot. Notice the arguments in the ordinate function; the index, Bray Curtis ("bray") or Jaccard ("jaccard"), is specified in the distance agurment, and with either of these metrics we will use method = "NMDS". The place were you might want to make some modifications is the k (this should be somewhere between 1-3), the maximum number of iterations (more than 100 will take several minutes to calcaulte), and the inital number of interations to try (50 in the example below). These can be modified to reduce the stress, with the goal of a number less than 0.20 (the lower the better). 
```{r}
physq_bc <- ordinate(physq_sub, 
                      method = "NMDS", 
                      k=2, maxit=100, try=50,
                      distance = "bray")
physq_bc
```

### Homework 8.3
Conduct an ordination analysis using Jaccard index. Record the stress value. 
```{r}


physq_j <- ordinate(physq_sub, 
                      method = "NMDS", 
                      k=2, maxit=100, try=50,
                      distance = "jaccard")
physq_j
```
#Stress value record

Square root transformation
Wisconsin double standardization
Run 0 stress 0.2330285 
Run 1 stress 0.2451595 
Run 2 stress 0.2360381 
Run 3 stress 0.2483864 
Run 4 stress 0.2566162 
Run 5 stress 0.2385795 
Run 6 stress 0.2440348 
Run 7 stress 0.2363355 
Run 8 stress 0.2545369 
Run 9 stress 0.2389295 
Run 10 stress 0.2441022 
Run 11 stress 0.239533 
Run 12 stress 0.2432584 
Run 13 stress 0.2324993 
... New best solution
... Procrustes: rmse 0.02124449  max resid 0.1643201 
Run 14 stress 0.2503801 
Run 15 stress 0.245271 
Run 16 stress 0.2594622 
Run 17 stress 0.2456399 
Run 18 stress 0.2426071 
Run 19 stress 0.2455424 
Run 20 stress 0.2512586 
Run 21 stress 0.2492041 
Run 22 stress 0.2417884 
Run 23 stress 0.2400289 
Run 24 stress 0.250031 
Run 25 stress 0.2497181 
Run 26 stress 0.4189509 
Run 27 stress 0.264684 
Run 28 stress 0.2610375 
Run 29 stress 0.2519296 
Run 30 stress 0.2546629 
Run 31 stress 0.2513913 
Run 32 stress 0.2652179 
Run 33 stress 0.2385338 
Run 34 stress 0.2595263 
Run 35 stress 0.2324081 
... New best solution
... Procrustes: rmse 0.01722926  max resid 0.1331893 
Run 36 stress 0.4190199 
Run 37 stress 0.2591236 
Run 38 stress 0.2489007 
Run 39 stress 0.2374388 
Run 40 stress 0.2439789 
Run 41 stress 0.2364555 
Run 42 stress 0.2356654 
Run 43 stress 0.2349675 
Run 44 stress 0.2500361 
Run 45 stress 0.243404 
Run 46 stress 0.2469292 
Run 47 stress 0.2440811 
Run 48 stress 0.2405152 
Run 49 stress 0.2623118 
Run 50 stress 0.2507328 
*** No convergence -- monoMDS stopping criteria:
    42: no. of iterations >= maxit
     8: stress ratio > sratmax

Call:
metaMDS(comm = veganifyOTU(physeq), distance = distance, k = 2,      try = 50, maxit = 100) 

global Multidimensional Scaling using monoMDS

Data:     wisconsin(sqrt(veganifyOTU(physeq))) 
Distance: jaccard 

Dimensions: 2 
Stress:     0.2324081 
Stress type 1, weak ties
No convergent solutions - best solution after 50 tries
Scaling: centring, PC rotation, halfchange scaling 
Species: expanded scores based on ‘wisconsin(sqrt(veganifyOTU(physeq)))’ 

### Plot results

For these plots we will use the phyloseq function plot_ordination() instead of ggplot(). Grouping variables can we assigned to color and shape, here they are differnet (study week and semester) but sometimes you may want them to be the same. 
```{r}
ordplot1 <- plot_ordination(physeq = physq_sub, 
                     ordination = physq_bc, 
                     type = "samples", 
                     color = "semester", 
                     shape = "study_week")
print(ordplot1)
```

### Modifications of ordination plots

The phyloseq plot functions work nicely with ggplot layers. They can modified with ggplot functions you already know. The process to save these plots is the same, use save_plot(). 

See this website for color options: http://sape.inf.usi.ch/quick-reference/ggplot2/colour 
And this site for shape options: http://sape.inf.usi.ch/quick-reference/ggplot2/shape 
```{r}
ordplot2 <- ordplot1 +
  geom_point(size=3) + #increase point size
  scale_color_manual(name = "Semester",
                     values = c("coral3", 
                                "skyblue3", 
                                "goldenrod3"),
                     labels = c("Fall 2016",
                                "Winter 2016", 
                                "Winter 2018")) + #change default color palette 
  scale_shape_manual(name = "Study week",
                     values = c(7, 8),
                     labels = c("week 1", "week 3")) #change default shapes 
print(ordplot2)
```

### Homework 8.4
Plot the results of the Jaccard ordination. Use colors and shapes different than those in the figure above. Clean up axis labels where necessary. Save the plot to figures folder. 
```{r}
ordplot3 <- plot_ordination(physeq = physq_sub, 
                     ordination = physq_j, 
                     type = "samples", 
                     color = "semester", 
                     shape = "study_week")
print(ordplot3)

ordplot4 <- ordplot3 +
  geom_point(size=3) + #increase point size
  scale_color_manual(name = "Semester",
                     values = c("grey69", 
                                "green3", 
                                "tan2"),
                     labels = c("Fall 2016",
                                "Winter 2016", 
                                "Winter 2018")) + #change default color palette 
  scale_shape_manual(name = "Study week",
                     values = c(9, 10),
                     labels = c("week 1", "week 3")) #change default shapes 
print(ordplot4)

save_plot(filename = "~/Documents/UMich_Bio201_F19/Lab8/figures/Jaccard ordination.pdf",
          plot = ordplot4)
```

### Statistical test

If you observe a pattern (separation of points from different groups) in your ordination plot, we can determine if that is to a significant level with analysis of similarities using distance matrices (ANOSIM or ADONIS). Using the ordination object from the section above, we run the test with the adonis() function, which is part of the vegan package. 
```{r}
#start with same phyloseq object as above 
dat_bray <- phyloseq::distance(physq_sub, method = "bray") #calculate BC index, get distance matrix

sampledf <- physq_sub %>% 
  sample_data(.) %>% #extract sample data from phyloseq object 
  as(., "data.frame") #convert to data frame for adonis()

# run test
adn_res <- adonis(formula = dat_bray ~ semester * study_week, 
                  data = sampledf)

# view results 
print(adn_res)
```

Review the output from the adonis function, this format should be slightly familar to you now; we are interested in the R-squared (R2) and the p-value (Pr(>F)) columns. The p-value indicates the groups are statistically significant. The R-squared value can be interpretd as the effect size, it shows the percentage of variation explained by the grouping categories (here study week and semester). For this test, an R-squared value close to 1 indicaates dissimilarity between the groups, while an R value near 0 indicates no significant dissimilarity between the groups. 

In our results above, the p-values of 0.001 indicate that the groupings of samples by semester or by study week is statistically significant. The R value of 0.025 and 0.008 are fairly close to 0, indicating that the grouping of samples based by semester or study week are weak. When this occurs (low p and low R), the variation of the samples in the tested groups likely is influencing the results of the test. 

### Homework 8.5
Conduct an ADONIS test on a distance matrix calculated with the Jaccard index. Write your interpretations of the data below, be sure to explicity state the calculated R- and p-values. 
```{r}
dat_ja <- phyloseq::distance(physq_sub, method = "jaccard") 

sampledf <- physq_sub %>% 
  sample_data(.) %>% #extract sample data from phyloseq object 
  as(., "data.frame") #convert to data frame for adonis()

# run test
adn_2 <- adonis(formula = dat_ja ~ semester * study_week, 
                  data = sampledf)

# view results 
print(adn_2)
```
In our results above, the p-values of 0.001 indicate that the groupings of samples by semester or by study week is statistically significant.  The R2 value of 0.028 and 0.008 are fairly close to 0, indicating that the grouping of samples based by semester or study week are weak.

-----
end
