---
title: "Bio201 Final Project"
author: "Yingyue Ding","Kevin Huang","Sunyimeng Lei", "Clarissa Halim" 
date: "11/18/2019"
output: html_document
---

Research Question: 
How does the consumption of a prebiotic effect the gut microbiomal community(diversity, changes in taxon abundance)?

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "~/Documents/Bio201ThursPM/")
```

# Load packages
```{r Load packages, eval=TRUE, echo=TRUE, message=FALSE, warning=FALSE, results='hide'}
library(tidyverse)
library(readxl)
library(broom)
library(cowplot)
library(agricolae)
library(phyloseq); packageVersion("phyloseq")
library(vegan)
set.seed(7)
```

# Import data

### Sequence variants 
Import shared/ESV/OTU table and format for phyloseq. Phyloseq works by matching rownames across matrix objects (This is similar to how we matched columns in joins of data frames). A two-dimensional matrix in R looks just like a data frame, however R matrices can be three-dimensional. A three-dimensional object is created when multiple two-dimensional objects are combined (we'll work through this in the upcoming sections). 
```{r}
#convert txt to valid cvs, I have done this step, and upload "species_avg_shared.csv" to our reporsitory.

#species_avg_shared <- read_delim("C:/Users/tina5/OneDrive/桌面/Bio201ThursPM/raw_data/species_avg_shared.txt",
#                            delim = "\t", escape_double = FALSE, 
#                            trim_ws = TRUE, na=c("NA"),
#                           col_types = list()) %>%   #import txt
# rename(participant_id_week = Participant_ID) %>% 
#  select(-Semester)#don't need
#  species_avg_shared$participant_id_week = paste(species_avg_shared$participant_id_week,
#                                                 species_avg_shared$Study_week, 
#                                                 sep="_") #create a combined name
#write.csv(species_avg_shared,file="C:/Users/tina5/OneDrive/桌面/Bio201ThursPM/curated_data/species_avg_shared.csv",quote=F,row.names = F)


species_avg_shared<-read_delim("~/Documents/Bio201ThursPM/curated_data/species_avg_shared.csv",
                            delim = ",", escape_double = FALSE, 
                            trim_ws = TRUE, na=c("NA"),
                            col_types = list()) 
sv_table <- read_delim("~/Documents/Bio201ThursPM/curated_data/species_avg_shared.csv",
                            delim = ",", escape_double = FALSE, 
                            trim_ws = TRUE, na=c("NA"),
                            col_types = list())  %>% 
  mutate_if(is.numeric, round, digits = 0) %>%
  column_to_rownames(var = "participant_id_week") %>% 
  as.matrix() %>%   #convert data frame to matrix object
  otu_table(., taxa_are_rows = FALSE) #specify type of phyloseq object

```

### Taxa 
Import taxonomy table and format for phyloseq.
```{r}
taxa <- read_delim("~/Documents/Bio201ThursPM/curated_data/seqvar_classifications.csv",
                            delim = ",", escape_double = FALSE, 
                            trim_ws = TRUE, na=c("NA"),
                            col_types = list()) %>%
  column_to_rownames(var = "sv") %>% #sequence variants (OTUs) need to be made into row names 
  as.matrix() %>%
  tax_table(.)  #specify type of phyloseq object
```

### Participant data 
Import sample data (where a sample is an individual microbiome sample) and format for phyloseq.
```{r}
all_data_wkly <- read_delim("~/Documents/Bio201ThursPM/raw_data/all_data_wkly.txt", 
                            delim = "\t", escape_double = FALSE, 
                            trim_ws = TRUE, 
                            col_types = cols(Bristol_numeric = col_number(),
                                             Bristol_score = col_character(), 
                                             pH = col_number())) %>% 
  rename_all(tolower) %>% 
  mutate(participant_id_week = participant_id)
  
  
all_data_wkly$participant_id_week = paste(all_data_wkly$participant_id,
                                                 all_data_wkly$study_week, 
                                                 sep="_")  #match with "sv_table"

all_data_wkly_inner <- inner_join(x = all_data_wkly, y = species_avg_shared, 
                                  by = c( "participant_id_week")) %>% 
                      select( participant_id_week, study_week, semester, supplement_consumed,
                              quantity_compliant, frequency, ph_median, ph_mean, bristol_median,
                              bristol_mean, blood_glucose_median, blood_glucose_mean, acetate_median,
                              acetate_mean, butyrate_median ,butyrate_mean, propionate_median,
                              propionate_mean) %>% 
  
                      distinct(., participant_id_week, .keep_all = TRUE) %>% #remove duplicate participant ids
                      column_to_rownames(var = "participant_id_week") %>% #participant IDs need to be made into row names
                      sample_data(.) #specify type of phyloseq object

```


# Create phyloseq objects
Phyloseq is an R package that has a suite of functions that make microbial community anaylsis a bit simpler. https://joey711.github.io/phyloseq/index.html It requires the data to be structured in three critical files (1) sequence variant (or OTU) table that contains counts of each sequence; (2) Taxanomic information that contains classifications from Phylum to species for each sequence variant; (3) sample information, which is the participant information we have been working with so far this semester. We imported these three files in the section above, now we have to combine them into one object. 
```{r}
physq_obj <- phyloseq(sv_table, taxa, all_data_wkly_inner) #combine three phyloseq objects created above 

physq_obj #view dimentions of new object 
```

# Richness
There are numerous alpha diversity metrics, however we are going to focus on number richness, evenness, and Simpson for our analyses. We will calculate, plot, extract information about alpha diversity with the plot_richness() funciton. 

### Calculate richness
```{r}
# Caculate richness

richness_df <- physq_obj %>%
 subset_samples(supplement_consumed == "BRMPS")%>%
   subset_samples(frequency == "2xdaily") %>%
 subset_samples(quantity_compliant == "yes") %>%
subset_samples(study_week == "week1" | study_week == "week3") %>%#subset for weeks of interest
  estimate_richness(., split = TRUE,  measures = c("Observed")) %>% #calculate richness
  rownames_to_column(var = "participant_id_week") %>% #reverse phyloseq object formatting
  inner_join(all_data_wkly, by = "participant_id_week")%>%
select(participant_id, study_week, Observed)%>%
  rename(richness = Observed) 
 
  

# save richness data 
write_delim(richness_df, path = "~/Documents/Bio201ThursPM/curated_data/richness_df.txt", delim = "\t")


```

### Plot richness
```{r}
# Plot richness
rich1 <- physq_obj %>%
 subset_samples(supplement_consumed == "BRMPS")%>%
   subset_samples(frequency == "2xdaily") %>%
 subset_samples(quantity_compliant == "yes") %>%
subset_samples(study_week == "week1" | study_week == "week3") %>%
  plot_richness(., "study_week", measures = c("Observed")) +
  ylab("Richness (Observed ESVs)") + xlab(NULL)

rich2 <- rich1 +   
  geom_violin(aes(color = study_week)) + #add violin in color
  geom_jitter(aes(color = study_week)) +  #add individual points in color 
  theme(legend.position = "none")
rich2

# save plot
save_plot(filename = "~/Documents/Bio201ThursPM/figures/rich2.pdf",
          plot = rich2,
          nrow = 1, ncol = 2,
          base_aspect_ratio = 1.1) 

```

```{r}
richness_df %>% 
  group_by(study_week) %>%
summarise(sample_size = n())

richness_normal<-richness_df 

  shapiro.test(richness_normal$richness)
ggplot(richness_normal, aes(x=richness)) +
  geom_histogram() 
qqnorm(richness_normal$richness); qqline(richness_normal$richness)

richness_wk1<- richness_df%>%
  filter(study_week == "week1") 
richness_wk3 <- richness_df %>%
  filter(study_week == "week3") 
var.test(x = richness_wk1$richness, 
         y = richness_wk3$richness, 
         alternative = "two.sided")


richness_df2 <- inner_join(x = richness_wk1, y = richness_wk3,
                    by = c("participant_id")) %>% 
  rename(richness_wk1 = richness.x,
         richness_wk3 = richness.y) 
  

t.test(x = richness_df2$richness_wk1, 
       y = richness_df2$richness_wk3,
       alternative = "two.sided", paired = TRUE)
```



# Simpson 

To calculate Simpson diveristy index we repeat the process above for calcuating richness, the only thing we have to change is the `measures = c("Simpson")` argument for the plot_richness() and estimate_richness() functions. 

### Calculate simpson
```{r}
simpson_df <- physq_obj %>%
 subset_samples(supplement_consumed == "BRMPS")%>%
   subset_samples(frequency == "2xdaily") %>%
 subset_samples(quantity_compliant == "yes") %>%
subset_samples(study_week == "week1" | study_week == "week3") %>% #subset for weeks of interest
  estimate_richness(., split = TRUE,  measures = c("Simpson")) %>% #calculate simpson
 
rownames_to_column(var = "participant_id_week") %>% #reverse phyloseq object formatting
  inner_join(all_data_wkly, by = "participant_id_week")%>%
select(participant_id, study_week, Simpson)%>%
  rename(simpson = Simpson) 

# save Simpson data 
write_delim(simpson_df, path = "~/Documents/Bio201ThursPM/curated_data/simpson_df.txt", delim = "\t")

```

### Plot simpson 
```{r}
# Simpson 
simp1 <- physq_obj %>%
  subset_samples(supplement_consumed == "BRMPS")%>%
   subset_samples(frequency == "2xdaily") %>%
 subset_samples(quantity_compliant == "yes") %>%
subset_samples(study_week == "week1" | study_week == "week3") %>%
  plot_richness(., "study_week", measures = c("Simpson")) +
  ylab("Simpson 1-D") + xlab(NULL)
simp1$layers <- simp1$layers[-1] #drop geom_point 
simp2 <- simp1 +   
  geom_violin(aes(color = study_week)) + #add violin in color
  geom_jitter(aes(color = study_week)) +  #add individual points in color 
  theme(legend.position = "none")
simp2


# save plot
save_plot(filename = "~/Documents/Bio201ThursPM/figures/simp2.pdf",
          plot = rich2,
          nrow = 1, ncol = 2,
          base_aspect_ratio = 1.1) 

```
```{r}
simpson_df %>% 
  group_by(study_week) %>%
summarise(sample_size = n())

simpson_normal<-simpson_df 

  shapiro.test(simpson_normal$simpson)
ggplot(simpson_normal, aes(x=simpson)) +
  geom_histogram() 
qqnorm(simpson_normal$simpson); qqline(simpson_normal$simpson)

simpson_wk1<- simpson_df%>%
  filter(study_week == "week1") 
simpson_wk3 <- simpson_df %>%
  filter(study_week == "week3") 
var.test(x = simpson_wk1$simpson, 
         y = simpson_wk3$simpson, 
         alternative = "two.sided")


simpson_df2 <- inner_join(x = simpson_wk1, y = simpson_wk3,
                    by = c("participant_id")) %>% 
  rename(simpson_wk1 = simpson.x,
         simpson_wk3 = simpson.y) 
  

wilcox.test(x = simpson_df2$simpson_wk1, 
       y = simpson_df2$simpson_wk3,
       alternative = "two.sided", paired = TRUE)
```

# Evenness

To determine if a community is even or uneven, we use stacked bar plots as a relative measure. Before generating the plot, we have to transform our abundance values from sequence counts to a percentage.

### Data transformation
```{r}
physq_sub <- physq_obj %>%
  subset_samples(supplement_consumed == "BRMPS")%>%
   subset_samples(frequency == "2xdaily") %>%
 subset_samples(quantity_compliant == "yes") %>%
subset_samples(study_week == "week1" | study_week == "week3")
physq_sub

physq_long <- physq_sub %>%
  transform_sample_counts(function(x) {x/sum(x)} ) %>%
  psmelt() %>%
  rename_all(tolower) %>%
  select(-quantity_compliant, -starts_with("bristol"), -ph_mean, -ph_median)
```

### Generate plot 
Now we are ready to make our plot. The same prinicples apply, we are just using a new ggplot function, geom_col(). 
```{r}
# vector of color hex codes 
taxa_colors <- c('#ff7f7f','#33b5e5',
                '#ffbb33','#aa66cc',
                '#ff3298', '#99cc00',
                '#cc0000','#007299',
                '#cc6c00','#6b238e',
                '#b20058', '#436500',
                '#ff4444','#0099cc',
                '#ff8800','#9933cc',
                '#e50072', '#669900',
                '#7f0000','#004c66',
                '#995100','#3c1451',
                '#660033', '#2c4c00',
                '#8ed5f0','#ffb2b2',
                '#bc93d1','#ffe564',
                '#d2fe4c','#ff7fbf')

# generate plot
ggplot(data = physq_long, 
       aes(x = study_week, #specify the arguments for each axis
           y = abundance)) + #set fill to the taxa rank used for glom
  geom_col() + #plot type for bars 
  facet_grid(~semester) + #panel for each semester
  xlab(NULL) + 
  theme(axis.text.x = element_text(angle = 90,
                                   vjust = 0.5)) + #clean up x-axis labels 
  ylab("Relative Abundance")
```


# Beta diveristy

### Conduct ordination analysis

This step calculates the diveristy index (using Bray Curtis in the example below), and does the ranking process of the diveristy indices to generate the information necessary to create a two-dimensional plot. Notice the arguments in the ordinate function; the index, Bray Curtis ("bray") or Jaccard ("jaccard"), is specified in the distance agurment, and with either of these metrics we will use method = "NMDS". The place were you might want to make some modifications is the k (this should be somewhere between 1-3), the maximum number of iterations (more than 100 will take several minutes to calcaulte), and the inital number of interations to try (50 in the example below). These can be modified to reduce the stress, with the goal of a number less than 0.20 (the lower the better). 
```{r}
# Bray Curtis
physq_sub <- physq_obj %>%
  subset_samples(supplement_consumed == "BRMPS")%>%
   subset_samples(frequency == "2xdaily") %>%
 subset_samples(quantity_compliant == "yes") %>%
subset_samples(study_week == "week1" | study_week == "week3")
physq_sub

physq_bc <- ordinate(physq_sub, 
                      method = "NMDS", 
                      k=2, maxit=100, try=50,
                      distance = "bray")
physq_bc

```

```{r}
# Jaccard
physq_j <- ordinate(physq_sub, 
                      method = "NMDS", 
                      k=2, maxit=100, try=50,
                      distance = "jaccard")
physq_j

```

### Plot results and Modifications of ordination plots
```{r}
# Bray Curtis
ordplot1 <- plot_ordination(physeq = physq_sub, 
                     ordination = physq_bc, 
                     type = "samples", 
                     color = "semester", 
                     shape = "study_week")
print(ordplot1)

```

```{r}
# Jaccard
ordplot2 <- plot_ordination(physeq = physq_sub, 
                     ordination = physq_j, 
                     type = "samples", 
                     color = "semester", 
                     shape = "study_week")
print(ordplot2)


```

### Statistical test
```{r}
# run Bray Curtis test
dat_bray <- phyloseq::distance(physq_sub, method = "bray") 

sampledf <- physq_sub %>% 
  sample_data(.) %>% 
  as(., "data.frame") 


adn_res <- adonis(formula = dat_bray ~  study_week, 
                  data = sampledf)


print(adn_res)

# Conclusion

```

```{r}
# run Jaccard test
dat_ja <- phyloseq::distance(physq_sub, method = "jaccard") 

sampledf <- physq_sub %>% 
  sample_data(.) %>% 
  as(., "data.frame") 


adn_2 <- adonis(formula = dat_ja ~ semester * study_week, 
                  data = sampledf)


print(adn_2)


# Conclusion

```
-----
end
